### 2011-07-21 Paper/Paper.tex
Draft finished

Given a hypothetical perfect synthesizer, performing scores with computers is a trivial task. However \textit{expressively} performing music is not and much research has focussed on this issue. Such research is interesting from a scientific point of view as well as a practical point of view. Artificial intelligence, cognitive science and musicology may all benefit from a better understanding of expression in music.

@@

A rather practical approach to expressive music performance is to recreate it. A performance rendering system is a computer program that automatically creates an expressive, humanlike performance of a musical score. Performance rendering systems have always faced a problem of evaluation. It is debatable how much we should value correlation of generated performances with human performances of the same piece. A Chopin interpretation of Arthur Rubinstein can be radically different from a Chopin interpretation of Vladimir Horowitz, yet both can be seen valid interpretations. This problem has been one of the motivations for a yearly performance rendering event where performance rendering systems compete and are judged by an audience.

@@

Section \ref{sec:approach} will describe the system as a whole. What we mean by structure level expression and why we think it is important is clarified in section \ref{sec:structure}. Section \ref{sec:method} will describe every component of the system in detail.

@@

The YQX system, as the authors admit, tended to sometimes produce nervous sounding changes in expression. He presents two extensions that seek to generate smoother performances. The problem with these extensions is, as Widmer admits himself, that the increased smoothness comes at the expense of expressitivity. To counter this, he adds three explicit rules that he uses to postprocess the performances. We think the reason that Widmer stumbled upon this tradeoff between expressiveness and nervousness because the nervousness is inherent to the way performances are generated at note-level. The system simply misses one component of expression, namely structure level expression.

@@

In this thesis, we propose a structure based performance rendering (SBPR) system. The system presented here ignores note level expression. Instead we will try to predict only constituent level expression. The assumption is that this kind of expression really exists in performances and that is different and independent from note level structure. We think that a constituent level system also corresponds better to how actual human performers play music. A performance redering system that only predicts note level expression would have rather meaningless fluctuations in tempo and dynamics as it does not have a notion of constituent level expression.

@@

The system will be similar to YQX in a number of ways, but instead of predicting expression per note, it will predict expression per constituent. Every constituent will be played with consistend expression, the articulation, dynamics and tempo change only at constituent breaks.

@@

The delta framework is based on the intuition that differences between notes indicate splits between constituents. The higher the difference, the higher the level of split in the parse tree (where the root note is at the highest level). Van der Berg proposes a delta rule that converts a set of differences, or deltas, between notes into a parsetree following this intuition.

@@

A set of basic features, $\Phi$, is assigned to each note. These are: \textbf{onset}, \textbf{pitch}, \textbf{loudness} and \textbf{release} (called offset by Van den Berg). From these two other features can be derived: \textbf{duration} and \textbf{virtual duration}. Duration is defined as \textbf{release($n_i$)} - \textbf{onset($n_i$)} while virtual duration is defined as \textbf{onset($n_{i+1}$)} - \textbf{onset($n_i$)}.

@@

A recursive rule, called the delta rule can be used to translate a list of deltas into a grouping structure. The deltarule, DeltaRule($M_{ij}$), where $M_{ij}$ is the ordered list of notes that is to be analyzed and $A$ is the resulting analysis, is defined as \footnote{The version here is a slightly reformulated, although functionally eqivalent, version of Van den Berg's deltarule. See \cite{markwin} for his original version.}:
\begin{algorithm}
\begin{algorithmic}
\STATE{$D_{i+1,j} \leftarrow \Delta M_{ij}$}
\STATE{$A \leftarrow []$}
\STATE $m \leftarrow \textbf{ max}(D)$
\FOR{$\delta \textbf{ in } D$}
\IF{$\delta = m$}
\STATE $p = \textbf{ index of }\delta \textbf{ in } D$
\STATE $q = \textbf{ index of the next occurence of m in }D \textbf{ or } j$
\STATE $\textbf{append DeltaRule(}M_{pq}\textbf{) to } A$
\ENDIF
\ENDFOR
\RETURN{$A$}
\end{algorithmic}
\caption{The delta rule}
\label{deltarule}
\end{algorithm}

The delta rule converts a sequence of notes into a nested list structure that can easily be interpreted as a tree. The deltarule can `parse' a piece of music into a \textit{parse tree}.

It also possible to define higher order delta function (deltas of deltas). Second order deltafunction can be used to differentiate a group of ascending notes from a group of notes that have wildly varying pitches. Van den Berg notes that this generates ambigous grouping structures as for example a second order deltarule needs at least three notes to be specified. For three notes, three different grouping structures are possible: $[[n_1, n_2], n_3]$, $[n_1, n_2, n_3]]$ and $[n_1, [n_2, n_3]]$. Because we want to use a second order deltafunction for segmentation only, we choose to use only one interpretation: $[n_1, n_2], n_3]$. This choice is motivated by the intuition that if in a group of three notes the delta function in some feature between the first two notes is low and the delta function between the second and third note is high (thus resulting in a high second order deltafunction) we would want a constituent split to happen between the second and third note. Note again that segmentation, like structure, may often be ambigous and multiple interpretations may be right; for our purposes it is important that at least one of these right interpretations is chosen.

Having defined first and second order deltafunctions and the deltarule, we can parse a piece of music into a delta tree or parse tree. Since we have a set of six basic note features, we have a set of six interpretations(parse trees) available for a piece of music. Intuition tells us that if a particular group of notes is found in multiple parse trees, this group of notes may be eligible to be a constituent. Van den Berg captures this intuition in the form of a \textit{yield rule}. A node in the parse tree is said to `yield' a group of notes if the node recursively contains this group of notes. If two or more nodes in different trees yield the same group of notes they should be connected according to the yield rule.

@@

A recursive structure like the parse trees generated by the delta framework is the kind of structure that we would ideally want to attribute to music. However, because of the problems with interpreting different parsetrees and because deltatrees do not necessarily represent the kind of structure that human musical intuition would attribute to music, we chose not to use a hierarchical structure and instead settle for a segmentation, derrived from deltatrees.

The goal is to find a `safe' method to segment music, that is, a method that does not generate segmentations that clash too much with human musical intuition. For this purpose the parse trees of delta onset (inter-onset interval, or simply IOI) and delta pitch (pitch interval) seem most suited: sudden jumps in pitch or onset often correspond to phrase transitions. These two trees are displayed in figure \ref{fig:pitchonsettrees}.

\begin{figure}
\centering
\subfigure[Delta onset tree]{
\includegraphics[scale=0.4]{img/onset_sym}
\label{fig:donset}
}
\subfigure[Second order delta pitch tree]{
\includegraphics[scale=0.3]{img/second_order_pitch_sym}
\label{fig:dpitch}
}
\caption{Parse trees of the first eight bars of piano sonata KV331 III. (Turkisch March)by Mozart }
\label{fig:pitchonsettrees}
\end{figure}

In the trees in figure \ref{fig:pitchonsettrees} we can see that some nodes contain only subnodes, some nodes contain only notes and some nodes contain notes and subnodes. To translate the tree into a segmentation we use algorithm \ref{alg:segmentation} starting with the root node.
\begin{algorithm}
\caption{Segmentation}
\label{alg:segmentation}
\begin{enumerate}
\item If the node is a note, at it as a singleton constituent to the segmentation
\item Expand the node.
\item If the node contains only subnodes, or less than $x$ notes, start from step 1 for every subnode and -note.
\item If the node contains at least $x$ notes, add the notes recursively contained by the current note to the segmentation.
\end{enumerate}
\end{algorithm}

The value of $x$ represents the tolerance of singleton constituents. We need this parameter because sometimes very long notes cause top level splits in the onset tree. We have found that setting it to the number of notes recursively contained (yielded) by the current node divided by 16 works quite well for works by Mozart. Our segmentation of the first eight bars of Mozart's Turkish March using a delta onset tree can be found in figure \ref{fig:mozartsegments}.\footnote{Note that if you use the delta tree from figure  \ref{fig:donset} you will not get this segmentation. This segmentation was based on a onset delta tree over the entire piece instead of just the first eight bars}

@@

%Using this modification of the delta rule we can bring the delta onset tree back a shallow tree in which it is safe to choose groupings defined by the first level nodes.

For some music like Mozart, using only a delta onset tree works very well. However, some music, like Bach's inventionen uses almost no differentiation in IOI. Hence, a segmentation based solely on inter onset interval will not serve our purposes for this music.

To overcome this problem we will use pitch. While Bach's inventionen use almost no differentiation in IOI they can be segmented nonetheless. We will not make any attempt to combine pitch interval trees and IOI trees. Instead we take the segmentation given by the IOI tree and try to subdivide every segment using a pitch interval tree.

The sort of changes in pitch that indicate constituent transitions or not characterized very well by a first order pitch interval tree. For example a melody that has four notes that are chromatically ascending followed by four notes that have octave intervals between them should be segmented into two constituents where one contains the four ascending notes and another one contains the notes with octave intervals. A first order pitch interval tree puts the four ascending notes in one segment but puts the four octave interval notes in four separate segments. For this reason we use a second order pitch interval tree that does not suffer from this.

@@

We can now convert a piece of music into a serie of constituents. These constituents will be used to predict expression so we must be able to charaterize them in a way correlates with the way they are performed. Analogous to YQX we are looking for the \textit{context} of the constituent as well as some description of the constituent itself.

YQX uses a set of three score features: pitch interval, duration ratio and I-R arch. The pitch interval is simply the difference in pitch between the current note and the next note. The duration ratio is the logarithmic ratio of the duration of the current note and the duration of the next note.  The I-R arch is is the distance to the nearest point of closure, where closure is calculated from the Implication-Realization analysis.

@@

The two features above provide information about the constituent context: if they are both zero the constituent is apparently similar in mean pitch and mean duration. At note level there is not much to say about the current note besides the pitch and duration. However at constituent we would also like to say something about the constituent itself. For this purpose the \textit{constituent features} \textbf{mean delta pitch} and \textbf{mean delta duration} are used. These features say something about the amount change in pitch and the amount of change in note duration:
\begin{description}
\item[\textbf{Mean delta pitch}] The mean of all absolute pitch intervals within one constituent. 
\item[\textbf{Mean delta duration}] The mean of all absolute differences in duration of succeeding notes within one constituent
\end{description}

@@

The complete set of score features consists of the two context features and the two constituent features:

@@

We are going to define our own expression parameters in a similar way but let us first look at some issues with the definitions YQX uses. The definition of articulations seems to be awkward and inconsistend. Awkward because if a score note is not followed by a rest, the notated silence after it is zero, rendering articulation undefined. Inconsisted because all the other features use logarithmic ratios and we see no reason to not use a logarithmic ratio for articulation as well.

IOI ratio and loudness is defined relative to mean performance tempo and loudness. However, to capture micro expression in the form of small changes in onset relative to the beat, it seems more logical to define this feature relative to the local tempo, instead of relative to the global tempo. The same argument can be made for defining loudness relative to local loudness instead of global loudness. Structure level expression may help to define the concepts of local tempo and local loudness as we can simply take the mean tempo and loudness within one constituent and take this to be the local tempo and loudness.

@@

We can now reduce a piece of music, represented as a sequence of notes $N_{ij}$, of which we have a score and a performance, to a sequence of score feature vectors $F$ and expression parameter vectors $E$. First, we segment the score into constituents:

@@

If we parse a number of pieces is this manner we can use it to do some statistics. We can for example calculate the features likely-hood which is the conditional probability of a feature vector $f_i$ given an expression vector $e_i$.

@@

Where $c(x)$ is the number of occurences of $x$ in the corpus.
The expression transition probability is defined by the unigram count of $e_i$ divided by the bigram count of $e_{i-1}, e_i$.
\begin{equation}
\label{transition}
P(e_i|e_{i-1}) = \frac{c(e_{i-1}, e_i)}{c(e_{i-1})}
\end{equation}
where we make the simplification that expression in one constituent is only dependend on the expression in the previous constituent which is of course not true.

@@

For all four performances the system was trained with a discretization factor of five. In practice, this means that the constituent features (
\ref{sec:features}) were discretized into five bins and the context features into a slightly larger number of bins since their values can be negative as well. The expression features where discretized into ten bins, with a sensitivity of five. The bins had to be this large to recognise most of the score features in the score that was to be performed.

@@

The system that we proposed reduces the idea of structure to a segmentation of the score into constituents based on a structural analysis. For the structural analysis we used the delta framework\ref{markwin}, proposed by Markwin van den Berg. We noticed that a segmentation based on onset delta trees works well for some music but does not recognize all constituent transitions. To overcome this we suggested to use a second order pitch delta tree, but we ended up discarding this idea because it still produced awkward segmentations.  Our system would therefore not work very well on music that has a consistend rhythmic structure like some of Bach's music. In the end, an onset tree segmentation worked good enough to produce musically relevant segmentations given the the right genre of music.

>>>>

Given a hypothetical perfect synthesizer, performing scores with computers is a trivial task. However \textit{expressively} performing music is not and much research has focussed on this issue. A wide selection of scientific fields may have interest in such research: Artificial intelligence, cognitive science and musicology all benefit from a better understanding of expression in music.

@@

A rather practical approach to expressive music performance is to recreate it. A performance rendering system is a computer program that automatically creates an expressive, humanlike performance of a musical score. Performance rendering systems have always faced a problem of evaluation. It is debatable how much we should value correlation of generated performances with human performances of the same piece. A Chopin interpretation of Arthur Rubinstein can be radically different from a Chopin interpretation of Vladimir Horowitz, yet both are considered high quality interpretations. This problem has been one of the motivations for a yearly performance rendering event where performance rendering systems compete and are judged by an audience.

@@

What we mean by structure level expression and why we think it is important is clarified in section \ref{sec:structure}. Section \ref{sec:approach} will describe the system as a whole. Section \ref{sec:method} will describe every component of the system in detail. Practical issues are discussed in section \ref{sec:implementation}. Results are presented in section \ref{sec:results} and discussed in section \ref{sec:discussion}. Suggestions for improving the system and future research can be found in section \ref{sec:futurework}.

@@

The YQX system, as Flossman et al. admit, tended to sometimes produce nervous sounding changes in expression. They present two extensions tailored towards generating smoother performances. The problem with these extensions is, as Flossman et al. also admit himself, is that the increased smoothness comes at the expense of expressitivity. To compensate for this, three explicit rules are added to postprocess the performances. We think the reason that Flossman et al. stumbled upon this tradeoff between expressiveness and nervousness is that the nervousness is inherent to note level performance rendering. Note level systems simply are not suited to capture one component of expression, namely structure level expression.

@@

In this thesis, we propose a structure based performance rendering (SBPR) system. The system presented here ignores note level expression. Instead we will try to predict only constituent (or structure) level expression. The assumption is that this kind of expression really exists in performances and that is different and independent from note level structure. We think that a constituent level system also corresponds better to how actual human performers play music. 
%A performance redering system that only predicts note level expression would have rather meaningless fluctuations in tempo and dynamics as it does not have a notion of constituent level expression.

@@

The system will be similar to YQX in a number of ways, but with the crucial difference that expression will not be predicted per note, but per constituent. Every constituent will be played with consistend expression, the articulation, dynamics and tempo change only at constituent breaks.

@@

The delta framework is based on the intuition that differences between features of notes indicate splits between constituents. The higher the difference, the higher the level of split in the parse tree (where the root note is at the highest level). Van der Berg proposes a delta rule that converts a set of differences, or deltas, between notes into a parsetree following this intuition.

@@

A set of basic features, $\Phi$, is assigned to each note. These are: \textbf{onset}, \textbf{pitch}, \textbf{loudness} and \textbf{release} (called offset by Van den Berg). From these, two other features can be derived: \textbf{duration} and \textbf{virtual duration}. Duration is defined as \textbf{release($n_i$)} - \textbf{onset($n_i$)} while virtual duration is defined as \textbf{onset($n_{i+1}$)} - \textbf{onset($n_i$)}.

@@

A recursive rule, called the delta rule can be used to translate a list of deltas into a grouping structure. This ruled, called the delta rule is shown in algorithm \ref{alg:deltarule}, where DeltaRule($M_{ij}$) is a recursive call to the algorithm itself, $M_{ij}$ is the ordered list of notes to be analyzed and $A$ is the resulting analysis.\footnote{The version here is a slightly reformulated, although functionally equivalent, version of Van den Berg's deltarule. See \cite{markwin} for his original version.}:
\begin{algorithm}
\begin{algorithmic}
\STATE{$D_{i+1,j} \leftarrow \Delta M_{ij}$}
\STATE{$A \leftarrow []$}
\STATE $m \leftarrow \textbf{ max}(D)$
\FOR{$\delta \textbf{ in } D$}
\IF{$\delta = m$}
\STATE $p = \textbf{ index of }\delta \textbf{ in } D$
\STATE $q = \textbf{ index of the next occurence of m in }D \textbf{ or } j$
\STATE $\textbf{append DeltaRule(}M_{pq}\textbf{) to } A$
\ENDIF
\ENDFOR
\RETURN{$A$}
\end{algorithmic}
\caption{The delta rule}
\label{alg:deltarule}
\end{algorithm}

The delta rule converts a sequence of notes into a nested list structure that can be interpreted as a tree. The deltarule can `parse' a piece of music into a \textit{parse tree}.

It is also possible to define higher order delta function (deltas of deltas). Second order deltafunction can be used to differentiate a group of ascending notes from a group of notes that have wildly varying pitches. Van den Berg notes that this generates ambigous grouping structures as for example a second order deltarule needs at least three notes to be specified. For three notes, three different grouping structures are possible: $[[n_1, n_2], n_3]$, $[n_1, n_2, n_3]]$ and $[n_1, [n_2, n_3]]$. Because we want to use a second order deltafunction for segmentation only, we choose to use only one interpretation: $[[n_1, n_2], n_3]$. This choice is motivated by the intuition that if in a group of three notes the delta function in some feature between the first two notes is low and the delta function between the second and third note is high (thus resulting in a high second order deltafunction) we would want a constituent split to happen between the second and third note. Note again that segmentation, like structure, may often be ambigous and multiple interpretations may be right; for our purposes it is important that at least one of these right interpretations is chosen.

Having defined first and second order deltafunctions and the deltarule, we can use first and second order deltafunctions to parse a piece of music into a parse tree. Since we have a set of six basic note features, we have a set of six interpretations(parse trees) available for a piece of music. Intuition tells us that if a particular group of notes is found in multiple parse trees, this group of notes may be eligible to be a constituent. Van den Berg captures this intuition in the form of a \textit{yield rule}. A node in the parse tree is said to `yield' a group of notes if the node recursively contains this group of notes. If two or more nodes in different trees yield the same group of notes they should be connected according to the yield rule.

@@

A recursive structure like the parse trees generated by the delta framework is the kind of structure that we would ideally want to attribute to music. However, because of the problems with interpreting different parsetrees and because deltatrees do not necessarily represent the kind of structure that human musical intuition would attribute to music, we chose not to use a hierarchical structure and instead settle for a segmentation, which can be more reliably derrived from deltatrees.

The goal is to find a `safe' method to segment music, that is, a method that generates segmentations that do not clash with human musical intuition. For this purpose the parse trees of delta onset (inter-onset interval, or simply \textsc{IOI}) and delta pitch (pitch interval) seem most suited: sudden jumps in pitch or onset often correspond to phrase transitions. 
\begin{figure}
\centering
\subfigure[Delta onset tree]{
\includegraphics[scale=0.4]{img/onset_sym}
\label{fig:donset}
}
\subfigure[Second order delta pitch tree]{
\includegraphics[scale=0.3]{img/second_order_pitch_sym}
\label{fig:dpitch}
}
\caption{Parse trees of the first eight bars of piano sonata KV331 III. (Turkisch March)by Mozart }
\label{fig:pitchonsettrees}
\end{figure}

A delta onset tree and a second order delta pitch tree of the first eight bars of Mozart's Turkish March can be found in figure \ref{fig:pitchonsettrees}. In his figure we can see that some nodes contain only subnodes, some nodes contain only notes and some nodes contain notes and subnodes. To translate the tree into a segmentation we use algorithm \ref{alg:segmentation} starting with the root node.
\begin{algorithm}
\caption{Segmentation}
\label{alg:segmentation}
For some node $n$, do:
\begin{enumerate}
\item If $n$ is a note, add it as a singleton constituent to the segmentation
\item Expand $n$.
\item If $n$ contains only subnodes, or less than $x$ notes, start from step 1 for every subnode and -note.
\item If $n$ contains at least $x$ notes, add the notes recursively contained by the current note to the segmentation.
\end{enumerate}
\end{algorithm}

The value of $x$ represents the tolerance of singleton constituents. We need this parameter because sometimes very long notes cause top level splits in the onset tree. We have found that setting it to the number of notes recursively contained (yielded) by the current node divided by 16 works quite well for works by Mozart. Our segmentation of the first eight bars of Mozart's Turkish March using a delta onset tree can be found in figure \ref{fig:mozartsegments}.\footnote{Note that if the delta tree from figure  \ref{fig:donset} is used the segmentation will be different. This segmentation was based on a onset delta tree over the entire piece instead of just the first eight bars}

@@

%Using this modification of the delta rule we can bring the delta onset tree down to a shallow tree in which it is safe to choose groupings defined by the first level nodes.

For some music like Mozart's, using only a delta onset tree works very well. However, some music, like Bach's Inventionen uses almost no differentiation in IOI. Hence, a segmentation based solely on inter onset interval will not work well for these pieces.

To overcome this problem we will use pitch. While Bach's Inventionen use almost no differentiation in IOI they can be segmented nonetheless. We will not make any attempt to combine pitch interval trees and IOI trees. Instead we take the segmentation given by the IOI tree and try to subdivide every segment using a pitch interval tree.

The sort of changes in pitch that indicate constituent transitions are not characterized very well by a first order pitch interval tree. For example a melody that has four notes that are chromatically ascending followed by four notes that have octave intervals between them should be segmented into two constituents where one contains the four ascending notes and another one contains the notes with octave intervals. A first order pitch interval tree puts the four ascending notes in one segment but puts the four octave interval notes in four separate segments. A second order pitch interval puts the split in this case exactly where we would want it.

@@

We can now convert a piece of music into a series of constituents. These constituents will be used to predict expression so we must be able to charaterize them in a way that correlates with the way they are performed. Analogous to YQX we are looking for the \textit{context} of the constituent as well as some description of the constituent itself.

YQX uses a set of three score features: pitch interval, duration ratio and I-R arch. The pitch interval is simply the difference in pitch between the current note and the next note. The duration ratio is the logarithmic ratio of the duration of the current note and the duration of the next note.  The I-R arch is is the distance to the nearest point of closure, where closure is calculated from the Implication-Realization analysis \cite{narmour1990analysis}.

@@

The two features above provide information about the constituent context: if they are both zero the constituent is apparently similar in mean pitch and mean duration. At note level there is not much to say about the current note besides the pitch and duration. However at constituent level we would also like to say something about the constituent itself. For this purpose the \textit{constituent features} \textbf{mean delta pitch} and \textbf{mean delta duration} are used. These features say something about the amount change in pitch and the amount of change in note duration:
\begin{description}
\item[\textbf{Mean delta pitch}] The mean of all absolute pitch intervals within one constituent. 
\item[\textbf{Mean delta duration}] The mean of all absolute differences in duration of succeeding notes within one constituent
\end{description}

@@

The complete set of score features consists of the two context features and the two constituent features.

@@

We are going to define our own expression parameters in a similar way but let us first look at some issues with the definitions YQX uses. The definition of articulations seems to be awkward and inconsistend. Awkward because if a score note is not followed by a rest, the notated silence after it is zero, rendering articulation undefined. Inconsisted because all the other features use logarithmic ratios and we see no reason not to use a logarithmic ratio for articulation as well.

IOI ratio and loudness are defined relative to mean performance tempo and loudness. However, to capture micro expression in the form of small changes in onset relative to the beat, it seems more logical to define this feature relative to the local tempo, instead of relative to the global tempo. The same argument can be made for defining loudness relative to local loudness instead of global loudness. Structure level expression may help to define the concepts of local tempo and local loudness as we can simply take the mean tempo and loudness within one constituent and take this to be the local tempo and loudness.

@@

We can now reduce a piece of music, represented as a sequence of notes $N_{ij}$, of which we have a score and a performance, to a sequence of \textit{score feature vectors} $F$ and \textit{expression parameter vectors} $E$. First, we segment the score into constituents:

@@

We can construct a corpus by parsing a number of works in this manner. Given such a corpus, the rendering of a performance can be formulated as maximizing the probability: $P(E|F)$. We can estimate an approximation of this probability by calculating feature likelyhoods and transition probabilities. We approximate the features likely-hood, which is the conditional probability of a feature vector $f_i$ given an expression vector $e_i$, with the following probability:

@@

Where $c(x)$ is the number of occurences of $x$ in the corpus.
The expression transition probability is approximated by the unigram count of $e_i$ divided by the bigram count of $e_{i-1}, e_i$.
\begin{equation}
\label{transition}
P(e_i|e_{i-1}) = \frac{c(e_{i-1}, e_i)}{c(e_{i-1})}
\end{equation}
where we make the simplification that expression in one constituent is only dependend on the expression in the previous constituent which is of course not true.

@@

For all four performances the system was trained with a discretization factor of five. In practice, this means that the constituent features (see section \ref{sec:scorefeatures}) were discretized into five bins and the context features into a slightly larger number of bins since their values can be negative as well. The expression features where discretized into ten bins, with a sensitivity of five. The bins had to be this large to recognise most of the score features in the score that was to be performed.

@@

The system that we proposed reduces the idea of structure to a segmentation of the score into constituents based on a structural analysis. For the structural analysis we used the delta framework \cite{markwin}, proposed by Markwin van den Berg. We noticed that a segmentation based on onset delta trees works well for some music but does not recognize all constituent transitions. To overcome this we suggested to use a second order pitch delta tree, but we ended up discarding this idea because it still produced awkward segmentations.  Our system would therefore not work very well on music that has a consistend rhythmic structure like some of Bach's music. In the end, an onset tree segmentation worked good enough to produce musically relevant segmentations given the the right genre of music.
